---
title: "GARCH Volatility Modelling: Extending the AAPL Forecasting Analysis"
format: html
---

## Define the Series
```{r setup}
set.seed(2026)
library(readr)
library(dplyr)
library(rugarch)
library(FinTS)
library(here)

TICKER <- "AAPL"    
Start_date <- as.Date("2017-01-01")
End_date <- as.Date("2019-12-31")
train_frac <- 0.66

raw <- read_csv(
  here::here("data", "sp500_stocks.csv"),
  col_types = cols(date = col_date(format = "%Y-%m-%d"), symbol = col_character())
)

stock_window <- raw %>%
  filter(symbol == TICKER, date >= Start_date, date <= End_date) %>%
  arrange(date) %>%
  mutate(trading_day = row_number(),
         log_close = log(close),
         log_return = log_close - lag(log_close))


n_total <- nrow(stock_window)
n_train <- floor(n_total * train_frac)

train <- stock_window[1:n_train, ]
test <- stock_window[(n_train + 1):n_total, ]

returns_vec <- train$log_return[!is.na(train$log_return)]


```

## Explore (Recap)
This project uses the same AAPL log returns series established and explored in detail in Project 1 Financial Forecasting using ARIMA. Key findings from that EDA that are relevent to this porject include: log returns show near-zero linear autocorrelation but substantial excess kurtosis (5.92 vs. 3 for a normal distribution) and visually apparent volatility clustering, concentrated particularly around the Q4 2018 price drop. This motivated Project 1's residual diagnostics, which confirmed the volatitlity clustering statistically (squared-residual Ljung-Box, p = 5.18e-12) but found no mechanism within the AR/MA/ARIMA model class to represent it, giving us the starting point for this project: GARCH volatility modelling.

```{r explore-recap}
plot(returns_vec, type = "l", col = "darkred",
     main = paste(TICKER, "- Training Set Log Returns"),
     xlab = "Trading Day", ylab = "Log Return")
abline(h = 0, lty = 2, col = "grey40")
```

## Specify the Mean
```{r joint mean comparison}
mean_orders <- list("(0,0)" = c(0,0),
                    "(1,0)" = c(1,0),
                    "(0,1)" = c(0,1),
                    "(1,1)" = c(1,1))

joint_results <- data.frame()

for (nm in names(mean_orders)){
    spec <- ugarchspec(
        variance.model = list(model = "sGARCH", garchOrder = c(1,1)), mean.model = list(armaOrder = mean_orders[[nm]], include.mean = TRUE), distribution.model = "std"
        )
    
    fit <- tryCatch(ugarchfit(spec = spec, data = returns_vec, solver = "hybrid"), error = function(e) NULL)
    if (is.null(fit)) next
  ic <- infocriteria(fit)
  joint_results <- rbind(joint_results, data.frame(
    MeanOrder = nm, AIC = ic["Akaike",], BIC = ic["Bayes",]
  ))


}

print(joint_results)

```


## Obtain Innovations

```{r obtain innovations}
fit <- arima(returns_vec, order = c(0,0,0))
innovations <- residuals(fit)

cat("Mean estimate:", coef(fit), "\n")
cat("Number of innovations:", length(innovations), "\n")
```

## Check for ARCH Effects

```{r arch effects test}
arch_results <- ArchTest(innovations, lags = 12)
print(arch_results)

lb_squared <- Box.test(innovations^2, lag = 10, type = "Ljung-Box")
print(lb_squared)
```

## Specify the Variance

```{r specify variance}
arch1 <- ugarchspec(
    variance.model = list(model = "sGARCH", garchOrder = c(1,0)),
    mean.model = list(armaOrder = c(0,0), include.mean = T), 
    distribution.model = "std"
    )

garch11 <- ugarchspec(
    variance.model = list(model = "sGARCH", garchOrder = c(1,1)),
    mean.model = list(armaOrder = c(0,0), include.mean = T), 
    distribution.model = "std"
    )

fit_arch1 <- ugarchfit(spec = arch1, data = returns_vec, solver = "hybrid")
fit_garch11 <- ugarchfit(spec = garch11, data = returns_vec, solver = "hybrid")

show(fit_arch1)
show(fit_garch11)
```

## Estimate

### "optimal parameters" table goes here and interpretation of the results

## Diagnose

### exploring all tests performed and there results, including Ljung-Box, ARCH-LM, and residual diagnostics (interpretation of the results)


## Compare Models (Part A: Residual Problem / Part B: Interval Calibration)

```{r compare models}
garch11_residuals <- residuals(fit_garch11, standardize = TRUE)

lb_before <- Box.test(innovations^2, lag = 10, type = "Ljung-Box")
lb_after <- Box.test(garch11_residuals^2, lag = 10, type = "Ljung-Box")

comparison <- data.frame(
    Model = c("ARIMA(0,0,0) only (Project 1)", "GARCH(1,1) standardized residuals"),
    X_squared = c(lb_before$statistic, lb_after$statistic),
    p_value = c(lb_before$p.value, lb_after$p.value)
)

print(comparison)
```

```{r coverage-test}
returns_full <- stock_window$log_return[-1]
n_train <- length(returns_vec)
n_total <- length(returns_full)

origins <- seq(n_train, n_total - 1, by = 10)

results_coverage <- data.frame()

for (i in origins) {
    data_sub <- returns_full[1:i]
    actual <- returns_full[i + 1]

    fit_a <- tryCatch(arima(data_sub, order = c(0,0,0)), error = function(e) NULL)
    if (!is.null(fit_a)) {
        fc_a <- predict(fit_a, n.ahead = 1)
        arima_lower <- fc_a$pred[1] - 1.96 * fc_a$se[1]
        arima_upper <- fc_a$pred[1] + 1.96 * fc_a$se[1]
        arima_covered <- actual >= arima_lower & actual <= arima_upper
        arima_width <- arima_upper - arima_lower
    } else {
        arima_covered <- NA
        arima_width <- NA
    }

    spec_i <- ugarchspec(
        variance.model = list(model = "sGARCH", garchOrder = c(1,1)),
        mean.model = list(armaOrder = c(0,0), include.mean = TRUE),
        distribution.model = "std"
    )
    fit_g <- tryCatch(ugarchfit(spec = spec_i, data = data_sub, solver = "hybrid"), error = function(e) NULL)

    if (!is.null(fit_g)) {
        fc_g <- ugarchforecast(fit_g, n.ahead = 1)
        g_mean <- as.numeric(fitted(fc_g))
        g_sigma <- as.numeric(sigma(fc_g))
        df_t <- coef(fit_g)["shape"]
        tq <- qt(0.975, df = df_t)
        garch_lower <- g_mean - tq * g_sigma
        garch_upper <- g_mean + tq * g_sigma
        garch_covered <- actual >= garch_lower & actual <= garch_upper
        garch_width <- garch_upper - garch_lower
    } else {
        garch_covered <- NA
        garch_width <- NA
    }

    results_coverage <- rbind(results_coverage, data.frame(
        origin = i, arima_covered, arima_width, garch_covered, garch_width
    ))
}

coverage_summary <- data.frame(
    Model = c("ARIMA(0,0,0)", "GARCH(1,1)"),
    Coverage = c(mean(results_coverage$arima_covered, na.rm = TRUE),
                 mean(results_coverage$garch_covered, na.rm = TRUE)),
    Avg_Width = c(mean(results_coverage$arima_width, na.rm = TRUE),
                  mean(results_coverage$garch_width, na.rm = TRUE)),
    n_origins = c(sum(!is.na(results_coverage$arima_covered)),
                  sum(!is.na(results_coverage$garch_covered)))
)
print(coverage_summary)
```


## Forecast and Report

## Discussion
