---
title: "R Lab: Fitting GARCH Models"
date: 2026-09-16
---

## Setup

```{r setup}
data(Tbrate, package="Ecdat")
library(tseries)
library(fGarch)

# r = the 91-day treasury bill rate
# y = the log of real GDP
# pi = the inflation rate

Tbill = Tbrate[,1]
Del.Tbill = diff(Tbill)
```

## Problem 1

Plot both Tbill and Del.Tbill using time series and ACF plots, and perform
ADF and KPSS tests on both series to assess stationarity and heteroskedasticity.

```{r problem1-plots}
plot(Tbill)
plot(Del.Tbill)

acf(Tbill)
acf(Del.Tbill)

adf.test(Tbill)
kpss.test(Tbill)

adf.test(Del.Tbill)
kpss.test(Del.Tbill)
```

### Answer: Stationarity and heteroskedasticity of Tbill / Del.Tbill

**Stationarity**

Tbill (level):
  ADF: p = 0.6075 -> fail to reject H0 (unit root)
  KPSS: p <= 0.01 -> reject H0 (stationarity)
  Both tests agree: Tbill is NON-STATIONARY.
  Confirmed visually - its ACF decays very slowly (still ~0.45
  at lag 5-6), the signature of a highly persistent, near
  unit-root series rather than short-lived dependence.

Del.Tbill (first difference):
  ADF: p <= 0.01 -> reject H0 (unit root)
  KPSS: p = 0.1 -> fail to reject H0 (stationarity)
  Both tests agree: Del.Tbill is STATIONARY.
  Confirmed visually - its ACF drops sharply after lag 1, close
  to a white noise pattern, with almost no spikes beyond the
  significance bounds.

Conclusion: First differencing removes the stochastic trend in
Tbill, restoring stationarity in Del.Tbill.

**Heteroskedasticity**

The stationarity tests and the ACF of Del.Tbill only describe the
CONDITIONAL MEAN (linear correlation) - they say nothing about the
variance. A flat ACF is consistent with pure white noise, but also
with a series whose variance changes over time (e.g. GARCH), since
volatility clustering is a property of the conditional variance,
not the conditional mean.

Visually, Del.Tbill shows periods of small fluctuations interrupted
by bursts of much larger swings (e.g. around the early-1980s
Volcker-era rate hikes) - this is volatility clustering: large
changes tend to be followed by large changes (of either sign), and
small by small.

This is confirmed formally by looking at the ACF of the squared
series:

```{r problem1-squared-acf}
acf(Del.Tbill^2, main = "ACF of Squared Del.Tbill")
```

Significant autocorrelation here (despite the raw series' ACF
being flat) indicates conditional heteroskedasticity - i.e. ARCH
effects - and motivates fitting a GARCH model to Del.Tbill's
variance, even though no further AR/MA structure is needed for
its mean.

## Problem 2

Fit an ARMA/GARCH model, identify which model is being fit, and report the
parameter estimates.

```{r problem2-fit}
garch.model.DelTbill = garchFit(formula = ~arma(1,0) + garch(1,0), Del.Tbill)
summary(garch.model.DelTbill)
garch.model.DelTbill@fit$matcoef
```

### Model being fit

`garchFit(formula = ~arma(1,0) + garch(1,0), data = Del.Tbill)`

This fits an AR(1) model for the conditional mean and an
ARCH(1) model for the conditional variance, with Normal
(Gaussian) innovations.

Mean equation (AR(1)):
  y_t = mu + ar1 * y_{t-1} + eps_t

Variance equation (ARCH(1)):
  sigma_t^2 = omega + alpha1 * eps_{t-1}^2

Innovations:
  eps_t = sigma_t * z_t,   z_t ~ N(0,1)

where y_t = Del.Tbill, and eps_t is the residual from the
mean equation at time t.

### Parameter estimates

|           | Estimate | Std.Error | t value | p-value  | Sig. |
|-----------|----------|-----------|---------|----------|------|
| mu        | 0.08350  | 0.05391   | 1.549   | 0.1214   | ns   |
| ar1       | 0.24163  | 0.07280   | 3.319   | 0.0009   | ***  |
| omega     | 0.33816  | 0.06145   | 5.503   | 3.7e-08  | ***  |
| alpha1    | 0.83483  | 0.24295   | 3.436   | 0.0006   | ***  |

Fitted equations:
  y_t       = 0.0835 + 0.2416 * y_{t-1} + eps_t
  sigma_t^2 = 0.3382 + 0.8348 * eps_{t-1}^2

Note: mu is not statistically significant (p = 0.12), so there
is no strong evidence the mean of Del.Tbill differs from zero
once the AR(1) term is accounted for. ar1, omega and alpha1 are
all significant at the 0.1% level.

```{r problem2-residuals}
res = residuals(garch.model.DelTbill)
res_std = res / garch.model.DelTbill@sigma.t

par(mfrow=c(2,3))
plot(res)
acf(res)
acf(res^2)
plot(res_std)
acf(res_std)
acf(res_std^2)
```

## Problem 3

Describe what each residual diagnostic plot shows and what it tells us about
the fit of the model.

### acf(res)

ACF of the raw residuals from the mean (ARMA) equation, before
standardising by the conditional SD. Bars stay within the
significance bounds (aside from a minor poke at lag 7) -> no
significant linear autocorrelation remains. The mean equation
(AR(1)) is adequate.

### acf(res^2)

ACF of the SQUARED raw residuals. Squaring removes sign and keeps
magnitude, revealing dependence that a plain ACF can miss. Clear
spikes outside the bounds (notably lag 4 and lag 8) indicate
volatility clustering / ARCH effects in the raw residuals - this
is exactly what motivates fitting a GARCH-type variance model.

### acf(res_std^2)

ACF of the squared STANDARDISED residuals (res/sigma.t), i.e.
after dividing out the model's own fitted conditional SD. Bars
now stay within the bounds -> the fitted ARCH(1) variance
equation has successfully absorbed the volatility clustering seen
in res^2. Consistent with the non-significant Ljung-Box (R^2) and
LM ARCH test results in the model summary.

### garch.model.Tbill@sigma.t

Contains the fitted conditional standard deviation, sigma_t, at
each time point in the sample - the model's estimated volatility
path implied by sigma_t^2 = omega + alpha1*eps_{t-1}^2. Used to
construct res_std = res / sigma.t.

### plot(res_std)

Noteworthy: even after standardising, the spread of res_std still
appears to widen over the index, similar to the raw residuals.
This suggests the conditional variance model has not fully
absorbed all the time-varying volatility - likely because the
model was fit to the non-stationary Tbill level series rather
than Del.Tbill, leaking trend-like behaviour into what looks like
residual heteroskedasticity. Note this visual funnel pattern can
persist without necessarily showing up as significant
autocorrelation in the ACF-based tests above.

## Problem 4

Find an ARMA/GARCH model for del.log.Tbill = diff(log(Tbill)), and consider
any advantages of working with log-differences rather than raw differences.

```{r problem4-fit}
del.log.Tbill = diff(log(Tbill))

# Explore first, same workflow as before
par(mfrow=c(1,2))
plot(del.log.Tbill, main="Differenced log(Tbill)")
acf(del.log.Tbill, main="ACF: del.log.Tbill")

adf.test(del.log.Tbill)
kpss.test(del.log.Tbill)

# Fit ARMA/GARCH, same specification as before for comparability
garch.model.logTbill = garchFit(formula = ~arma(1,0) + garch(1,0), data = del.log.Tbill)
summary(garch.model.logTbill)
```

### ARMA(1,0)+ARCH(1) fit on del.log.Tbill = diff(log(Tbill))

Stationarity:
  ADF: p <= 0.01 -> reject unit root
  KPSS: p = 0.080 -> fail to reject stationarity (borderline but OK)
  Both agree: del.log.Tbill is stationary.

Fitted model:
  y_t       = 0.0119 + 0.4901 * y_{t-1} + eps_t
  sigma_t^2 = 0.01206 + 0.6206 * eps_{t-1}^2

mu is not significant (p=0.21); ar1, omega, alpha1 all significant.
ar1 (0.490) is notably higher than in the Del.Tbill fit (0.242),
suggesting log-differencing reveals more short-term mean
dependence - consistent with relative (%) changes being more
persistent period-to-period than raw point changes.

Diagnostics:
  - Ljung-Box on raw residuals: not significant -> mean eq. adequate
  - Ljung-Box on squared residuals / LM ARCH test: not significant
    overall (Q(10) borderline at p=0.069) -> ARCH(1) variance eq.
    largely adequate
  - Jarque-Bera / Shapiro-Wilk: still strongly reject normality,
    and MORE strongly than the Del.Tbill fit (JB stat 45.8 vs 27.0)
    -> heavy tails persist, Student-t innovations still worth trying

Note: AIC is not directly comparable to the Del.Tbill model's AIC,
since the two are fit to differently-scaled transformed data.

## 8. On Black Monday, the return on the S&P 500 was −22.8%. Ouch!

This exercise attempts to answer the question, "what was the conditional
probability of a return this small or smaller on Black Monday?"
"Conditional" means given the information available the previous trading
day.

```{r blackmonday-setup}
library(Ecdat)
library(fGarch)
data(SP500, package="Ecdat")
returnBlMon = SP500$r500[1805]
x = SP500$r500[(1804-2*253+1):1804]
plot(c(x,returnBlMon))
results = garchFit(~arma(1,0)+garch(1,1), data=x, cond.dist="std")
dfhat = as.numeric(results@fit$par[6])
forecast = predict(results, n.ahead=1)

z = (returnBlMon - forecast$meanForecast) / forecast$standardDeviation
p = pt(z, df = dfhat)

res_std = residuals(results, standardize = TRUE)

par(mfrow = c(2, 2))
plot(res_std, type = "l", main = "Standardized Residuals")
acf(res_std, main = "ACF: Standardized Residuals")
acf(res_std^2, main = "ACF: Squared Standardized Residuals")

Box.test(res_std, lag = 10, type = "Ljung-Box")
Box.test(res_std^2, lag = 10, type = "Ljung-Box")
```

### Q2 answer

All three diagnostics (time series plot, ACF of standardized
residuals, ACF of squared standardized residuals) show no
remaining structure - residuals sit within significance bounds,
and Ljung-Box tests are non-significant for both the raw (p=0.92)
and squared (p=0.89) standardized residuals.

Conclusion: the AR(1)/GARCH(1,1)-t model fits adequately - both
the conditional mean and conditional variance are well specified
over the 506-day training window. This says nothing about Black
Monday itself, which was excluded from the fit; the model's
adequacy here is consistent with Black Monday being an extreme,
essentially unpredictable event relative to this fitted model.

```{r blackmonday-arch1}
results_arch1 = garchFit(~arma(1,0) + garch(1,0), data = x, cond.dist = "std")
summary(results_arch1)

# summary()'s reported AIC is normalized (divided by n) - multiply
# back up by sample size to get comparable AIC values.
n = length(x)
AIC_garch11 = results@fit$ics["AIC"] * n
AIC_arch1   = results_arch1@fit$ics["AIC"] * n
c(GARCH11 = AIC_garch11, ARCH1 = AIC_arch1)

res_std_arch1 = residuals(results_arch1, standardize = TRUE)
par(mfrow = c(1, 2))
acf(res_std_arch1, main = "ACF: Std. Residuals, ARCH(1)")
acf(res_std_arch1^2, main = "ACF: Squared Std. Residuals, ARCH(1)")

Box.test(res_std_arch1^2, lag = 10, type = "Ljung-Box")
```

### Q3 answer

AIC(GARCH11) = -3298.428, AIC(ARCH1) = -3298.121 - nearly identical,
GARCH(1,1) only marginally better.

Residual diagnostics for AR(1)/ARCH(1)-t: Ljung-Box on squared std.
residuals (p=0.27,0.43,0.66) and LM ARCH test (p=0.15) are all
non-significant, and the ACF plots show no structure beyond lag 0
-> formally, the ARCH(1) model also appears adequate.

However: alpha1 is NOT statistically significant in this fit
(p=0.375), unlike in the GARCH(1,1) model where alpha1 was clearly
significant (p=0.003). This suggests the ARCH(1) model's apparent
adequacy is driven mainly by its very heavy-tailed Student-t
innovations (shape ~ 3.96) rather than genuine time-varying
conditional variance - i.e. the model explains extreme returns via
fat tails rather than volatility clustering.

Conclusion: AR(1)/ARCH(1)-t is statistically adequate by these
diagnostics, but GARCH(1,1)-t remains the more defensible model,
since it achieves the same (or marginally better) fit with all
coefficients - including the ARCH term - individually significant
and interpretable.

```{r blackmonday-ar1}
fit_ar1 = arima(x, order = c(1, 0, 0))
summary(fit_ar1)

res_ar1 = residuals(fit_ar1)
res_ar1_std = res_ar1 / sd(res_ar1)   # standardize by constant SD (no time-varying sigma)

par(mfrow = c(1, 2))
acf(res_ar1_std, main = "ACF: AR(1) Residuals")
acf(res_ar1_std^2, main = "ACF: Squared AR(1) Residuals")

Box.test(res_ar1_std, lag = 10, type = "Ljung-Box")
Box.test(res_ar1_std^2, lag = 10, type = "Ljung-Box")
```

### Answer

Raw residual diagnostics:
  Ljung-Box, Q(10): p = 0.922 -> no significant linear
  autocorrelation remains; the AR(1) mean equation is adequate,
  essentially matching the result from the GARCH(1,1) fit.

Squared residual diagnostics:
  Ljung-Box, Q(10): p = 0.142 -> technically fails to reject at
  the 5% level, but the test statistic (14.74) is roughly 3x
  larger than the GARCH(1,1) model's (5.04) on the same data -
  noticeably weaker evidence of adequacy than either GARCH-type
  fit.

Conclusion: A Gaussian AR(1) model has no variance equation at
all, so it has no mechanism to capture volatility clustering.
While the squared-residual test does not formally reject at lag
10, the elevated test statistic points toward remaining ARCH
structure that a constant-variance model cannot account for. This
matches the stylised facts of return series (near-zero
autocorrelation in levels, strong autocorrelation in squares) and
suggests the Gaussian AR(1) model is, at best, borderline and
likely NOT adequate - unlike the GARCH(1,1) and ARCH(1) models,
which both showed comfortably non-significant squared-residual
autocorrelation across all lags tested.
