---
title: "Financial Forecasting: AR/MA/ARIMA Model Comparison on AAPL - FRNBEN003"
format: html
---
## Forecast Horizon and Justification

> **Draft — placeholder to be revised once results are in.** 

This project forecasts AAPL daily closing price and log returns over three
horizons — 1 trading day, 5 trading days (one week), and 20 trading days
(approximately one month) — rather than a single fixed horizon. This choice is
deliberate rather than a default.

Financial return series are widely documented to behave close to a random walk:
returns show near-zero linear autocorrelation at almost any lag, meaning today's
return carries very little information about tomorrow's direction. This has a
direct consequence for forecast horizon selection: the *shortest* horizon is
where an AR/MA/ARIMA model has the best chance of adding value beyond a naive
benchmark (e.g. "tomorrow's price equals today's price"), since any genuine
short-range dependence in the series — however weak — is most detectable and
most exploitable at short lags. As the horizon lengthens, the series' near-random-
walk character means forecast uncertainty compounds quickly and any linear
structure the model captured at short lags contributes proportionally less, so
accuracy is expected to degrade — this project treats that expected degradation
as a finding to demonstrate empirically, not something to avoid by only
forecasting one horizon.

Practically, near-term forecasting (1-day) reflects genuine short-horizon use
cases (e.g. next-session positioning), while the medium horizons (5-day, 20-day)
test whether the fitted model retains any predictive value beyond the immediate
next step, or whether its performance collapses toward a naive benchmark — a
meaningful comparison given the efficient-market expectation that little linear
structure should persist at longer horizons. Comparing performance (RMSE/MAE)
across all three horizons, rather than reporting a single number, therefore
provides a more complete and more honest picture of the model's actual
forecasting value than a single-horizon result would.

<!--
TODO once results are in:
- Swap in the actual finding (did accuracy degrade as expected, or was there
  a surprise?)
- Reference the specific AR/MA/ARIMA candidates by name/order once chosen
- Tie the horizon discussion to both price and returns series explicitly if
  the comparison ends up being a major part of the write-up
-->


## EDA

```{r setup}
set.seed(2026)
library(readr)
library(dplyr)
library(tsibble)
library(feasts)
library(ggplot2)
library(here)
```

```{r load data}
TICKER <- "AAPL"
start_date <- as.Date("2017-01-01")
end_date <- as.Date("2019-12-31")
Train_frac <- 0.66

raw <- read_csv(
  here::here("data", "sp500_stocks.csv"),
  col_types = cols(
    date   = col_date(format = "%Y-%m-%d"),
    open   = col_double(),
    high   = col_double(),
    low    = col_double(),
    close  = col_double(),
    volume = col_double(),
    symbol = col_character()
  )
)

stock_window <- raw %>%
  filter(symbol == TICKER,
         date >= start_date,
         date <= end_date) %>%
  arrange(date) %>% 
  mutate(trading_day = row_number()) 

n_total <- nrow(stock_window)
n_train <- floor(n_total * Train_frac)

train <- stock_window %>% filter(trading_day <= n_train)
test <- stock_window %>% filter(trading_day > n_train)

cat("Total obs:", n_total, "\n")
cat("Train:", nrow(train), "|", as.character(min(train$date)), "to", as.character(max(train$date)), "\n")
cat("Test: ", nrow(test),  "|", as.character(min(test$date)),  "to", as.character(max(test$date)), "\n")

```

```{r eda-plots}
train <- train %>%
  mutate(log_close = log(close),
         log_return = log_close - lag(log_close))

# Price Plot
p_price <- ggplot(train, aes(x = date, y = close)) +
    geom_line(color = "steelblue") +
    labs(title = paste(TICKER, "- Training Set Closing Price"),
    x = "Date", y = "Close Price ($)") +
    scale_x_date(date_breaks = "3 months", date_labels = "%b %Y")


p_price

p_returns <- ggplot(train %>% filter(!is.na(log_return)), aes(x = date, y = log_return)) +
    geom_line(color = "darkorange", linewidth = 0.3) +
    geom_hline(yintercept = 0, color = "grey40", linetype = "dashed") +
    labs(title = paste(TICKER, "- Training Set Log Returns"),
    x = "Date", y = "Log Return") +
    scale_x_date(date_breaks = "3 months", date_labels = "%b %Y")
p_returns

```

```{r eda-distribution}
returns_clean <- train %>% filter(!is.na(log_return))

summary_stats <- data.frame(
    Series = c("Close Price", "Log Return"),
    Mean = c(mean(train$close), mean(returns_clean$log_return)),
    SD = c(sd(train$close), sd(returns_clean$log_return)),
    Skewness = c(
        mean((train$close - mean(train$close))^3) / sd(train$close)^3,
        mean((returns_clean$log_return - mean(returns_clean$log_return))^3) / sd(returns_clean$log_return)^3
    ),
    Kurtosis = c(
        mean((train$close - mean(train$close))^4) / sd(train$close)^4,
        mean((returns_clean$log_return - mean(returns_clean$log_return))^4) / sd(returns_clean$log_return)^4
    )
)

print(summary_stats)

# Histogram of returns vs normal
Hist_returns <- ggplot(returns_clean, aes(x = log_return)) +
    geom_histogram(aes(y = after_stat(density)), bins = 40, fill = "steelblue", alpha = 0.6) +
    stat_function(fun = dnorm, args = list(mean = mean(returns_clean$log_return), sd = sd(returns_clean$log_return)), color = "red", linewidth = 1) +
    labs(title = paste(TICKER, "- Log Returns Distribution vs Normal"),
         x = "Log Return", y = "Density")
Hist_returns
#QQ - plot
QQ_returns <- ggplot(returns_clean, aes(sample = log_return)) +
    stat_qq() +
    stat_qq_line(color = "red") +
    labs(title = paste(TICKER, "- Log Returns Q-Q Plot"),
         x = "Theoretical Quantiles", y = "Sample Quantiles")
QQ_returns
```

```{r eda-acf}
returns_vec <- returns_clean$log_return

par(mfrow = c(3, 2))

acf(train$close, main = "ACF: Price", lag.max = 40)
pacf(train$close, main = "PACF: Price", lag.max = 40)

acf(returns_vec, main = "ACF: Returns", lag.max = 40)
pacf(returns_vec, main = "PACF: Returns", lag.max = 40)

acf(returns_vec^2, main = "ACF: Squared Returns", lag.max = 40)
pacf(returns_vec^2, main = "PACF: Squared Returns", lag.max = 40)

par(mfrow = c(1, 1))

```

## Stationarity
```{r stationarity-test}
library(tseries)

adf_price <- adf.test(train$close)
kpps_price <- kpss.test(train$close, null = "Level")

adf_returns <- adf.test(returns_vec)
kpps_returns <- kpss.test(returns_vec, null = "Level")

stationarity_results <- data.frame(
    Series = c("Close Price", "Close Price", "Log Return", "Log Return"),
    Test = c("ADF", "KPSS", "ADF", "KPSS"),
    Statistic = c(adf_price$statistic, kpps_price$statistic, adf_returns$statistic, kpps_returns$statistic),
    P_Value = c(adf_price$p.value, kpps_price$p.value, adf_returns$p.value, kpps_returns$p.value)
)

print(stationarity_results)


```

## Stationarity

Formal stationarity testing was carried out using the ADF and KPSS tests, which
have opposite null hypotheses and are read together for a more reliable
conclusion (ADF's null is a unit root/non-stationarity; KPSS's null is
stationarity).

| Series | Test | Statistic | p-value |
|---|---|---|---|
| Close Price | ADF | -0.793 | 0.962 |
| Close Price | KPSS | 6.856 | 0.010 |
| Log Return | ADF | -7.565 | 0.010 |
| Log Return | KPSS | 0.551 | 0.030 |

**Price**: ADF fails to reject the unit-root null (p = 0.962) and KPSS rejects
stationarity (p = 0.010). Both tests agree: price is non-stationary. This
confirms the visual evidence from the EDA (persistent upward trend, ACF
decaying extremely slowly even at lag 40) and the PACF's sharp cutoff after
lag 1, consistent with a random-walk-type process. Price requires differencing
(d = 1) before AR/MA-type modelling.

**Log returns**: ADF strongly rejects the unit-root null (p = 0.010), while
KPSS's rejection (p = 0.030) is borderline and most plausibly attributable to
the conditional heteroskedasticity (volatility clustering) already identified
in the EDA — via the widened swings visible in the returns plot, excess
kurtosis (5.92), and the more persistent squared-returns ACF — rather than
genuine non-stationarity in the mean. KPSS is known to be sensitive to this
kind of heteroskedasticity even when the mean is stationary. Log returns are
therefore treated as stationary (d = 0) for AR/MA/ARMA modelling.

**Decision going forward:** price will be modelled via ARIMA(p,1,q) (differencing
handled internally); log returns will be modelled via ARMA(p,q) (already
stationary, no further differencing). Because these two branches use different
orders of differencing, AICc cannot be used to compare across them directly —
only within each branch. Cross-branch comparison (price-based vs returns-based
forecasts) will instead rely on out-of-sample RMSE from the test set, which
remains valid regardless of differencing order.

<!--
TODO: revisit this section once order identification and model fitting are
done - check this interpretation still holds and reference the final chosen
models here if useful.
-->

## ACF/PACF Identification
``` {r order-id-acf-pacf}
par(mfrow = c(1, 2))
acf(returns_vec, main = "ACF: Log Returns (training)", lag.max = 20)
pacf(returns_vec, main = "PACF: Log Returns (training)", lag.max = 20)

par(mfrow = c(1, 1))

n <- length(returns_vec)
ci <- 1.96/sqrt(n)
acf_vals <- acf(returns_vec, lag.max = 20, plot = FALSE)
pacf_vals <- pacf(returns_vec, lag.max = 20, plot = FALSE)

cat("95% significance bounds: +/-", round(ci,4), "\n")
cat("Significant ACF lags (excl. lag 0):", which(abs(acf_vals$acf[-1]) > ci), "\n")
cat("Significant PACF lags:", which(abs(pacf_vals$acf) > ci), "\n")


diff_price <- diff(train$close)

par(mfrow = c(1, 2))
acf(diff_price, main = "ACF: Differenced Price (training)", lag.max = 20)
pacf(diff_price, main = "PACF: Differenced Price (training)", lag.max = 20)
par(mfrow = c(1, 1))

n_d <- length(diff_price)
ci_d <- 1.96/sqrt(n_d)
acf_vals_d <- acf(diff_price, lag.max = 20, plot = FALSE)
pacf_vals_d <- pacf(diff_price, lag.max = 20, plot = FALSE)

cat("95% significance bounds (diff price): +/-", round(ci_d,4), "\n")
cat("Significant ACF lags (excl. lag 0):", which(abs(acf_vals_d$acf[-1]) > ci_d), "\n")
cat("Significant PACF lags (diff price):", which(abs(pacf_vals_d$acf) > ci_d), "\n")
```

## ACF/PACF Identification

> **Draft — placeholder to be revised once results are in.**

Order identification was carried out on the returns series only (already
established as stationary), using ACF and PACF computed to lag 20, with
significance assessed numerically against the ±1.96/√n critical bound rather
than by visual inspection alone.

Lags 1–6 fall comfortably within the significance bounds in both ACF and PACF,
showing no early cutoff or gradual decay pattern — consistent with the
near-market-efficiency expected of daily equity returns (little to no linear
autocorrelation at short lags). Isolated significant spikes appear at lags 7,
8, 10, and 17, flagged identically in both ACF and PACF. This does not match
either the textbook AR signature (PACF cutoff, ACF decay) or MA signature (ACF
cutoff, PACF decay), and is most plausibly explained by multiple-testing noise
(with 20 lags tested at a 95% threshold, roughly 1 spurious spike is expected
by chance) rather than genuine structure, particularly as the flagged lags
have no clear interpretable spacing (e.g. no weekly/5-day pattern).

**Decision:** given the weak and ambiguous short-lag signal, small candidate
orders are fitted and compared via AICc and residual diagnostics, rather than
reading a specific (p, q) directly off the plot. This is a deliberate choice
given the data, not a shortcut — forcing a larger/more specific order onto an
ambiguous ACF/PACF read would not be justifiable.

**Candidates — Returns branch (ARMA, d = 0):**

- ARIMA(0,0,0) — white noise / naive baseline
- AR(1)
- MA(1)
- ARMA(1,1)

These will be compared via AICc within this branch, and checked with residual
diagnostics (Ljung-Box), before moving on to out-of-sample evaluation.

**Price branch identification.** Order identification was repeated on the
first-differenced raw close price (not log price, to keep this branch
conceptually distinct from the returns branch rather than redundant with it -
differencing log price would be mathematically equivalent to log returns).

The differenced price ACF/PACF show the same pattern as the returns branch:
lags 1-6 within the significance bounds in both plots, with isolated
significant spikes at lags 7, 8, 10, and 17, flagged identically in ACF and
PACF (95% bound = ±0.088). This near-identical structure between the two
branches is itself a notable finding: despite deliberately using raw price
differences rather than log returns, the two series behave very similarly.
This is consistent with the standard result that log returns closely
approximate percentage price changes when period-to-period moves are small
relative to the price level, as is typical for daily equity data - the
distinction between "dollar change" and "log/percentage change" matters much
less when moves are a small fraction of price.

**Candidates - Price branch (ARIMA, d = 1):**

- ARIMA(0,1,0) - random walk baseline
- ARIMA(1,1,0)
- ARIMA(0,1,1)
- ARIMA(1,1,1)

Given the similarity in ACF/PACF structure to the returns branch, the same
small-candidate rationale applies here for the same reason: no clean AR or MA
signature to justify a larger specific order.

<!--
TODO once fitting is done (both branches):
- Report actual AICc tables
- Note which candidate won each branch and why
- Revisit whether the lag 7/8/10/17 spikes deserve more discussion in light of
  final residual diagnostics
- Consider whether the price/returns similarity finding changes how the final
  cross-branch RMSE comparison should be framed/discussed
-->

## Model Fitting

```{r fit-returns}
aicc <- function(fit, n_eff){
    k <- length(fit$coef) + 1 #+1 for sigma^2
    fit$aic + (2 * k * (k+1))/(n_eff - k - 1)

}

fit_r_000 <- arima(returns_vec, order = c(0, 0, 0))
fit_r_100 <- arima(returns_vec, order = c(1, 0, 0))
fit_r_001 <- arima(returns_vec, order = c(0, 0, 1))
fit_r_101 <- arima(returns_vec, order = c(1, 0, 1))

n_r <- length(returns_vec)
returns_results <- data.frame(
    Model = c("ARIMA(0,0,0)", "AR(1)", "MA(1)", "ARMA(1,1)"),
    AICc = c(aicc(fit_r_000, n_r), aicc(fit_r_100, n_r), aicc(fit_r_001, n_r), aicc(fit_r_101, n_r))
)

print(returns_results)
```

```{r fit-price}
fit_p_010 <- arima(train$close, order = c(0, 1, 0))
fit_p_110 <- arima(train$close, order = c(1, 1, 0))
fit_p_011 <- arima(train$close, order = c(0, 1, 1))
fit_p_111 <- arima(train$close, order = c(1, 1, 1))

n_p <- length(train$close) - 1
price_results <- data.frame(
    Model = c("ARIMA(0,1,0)", "ARIMA(1,1,0)", "ARIMA(0,1,1)", "ARIMA(1,1,1)"),
    AICc = c(aicc(fit_p_010, n_p), aicc(fit_p_110, n_p), aicc(fit_p_011, n_p), aicc(fit_p_111, n_p))
)

print(price_results)
```


```{r residual-diagnostics}
lb_returns <- Box.test(residuals(fit_r_000), lag = 20, type = "Ljung-Box", fitdf = 0)
lb_price <- Box.test(residuals(fit_p_010), lag = 20, type = "Ljung-Box", fitdf = 0)

cat("Returns model = ARIMA(0,0,0):\n")
print(lb_returns)
cat("\nPrice model = ARIMA(0,1,0):\n")
print(lb_price)
```

```{r residual-diagnostics-compare}
lb_returns_10 <- Box.test(residuals(fit_r_000), lag = 10, type = "Ljung-Box", fitdf = 0)
lb_returns_20 <- Box.test(residuals(fit_r_000), lag = 20, type = "Ljung-Box", fitdf = 0)
lb_price_10   <- Box.test(residuals(fit_p_010), lag = 10, type = "Ljung-Box", fitdf = 0)
lb_price_20   <- Box.test(residuals(fit_p_010), lag = 20, type = "Ljung-Box", fitdf = 0)

ljung_box_summary <- data.frame(
  Model = rep(c("Returns ARIMA(0,0,0)", "Price ARIMA(0,1,0)"), each = 2),
  Lag = c(10, 20, 10, 20),
  p_value = c(lb_returns_10$p.value, lb_returns_20$p.value,
              lb_price_10$p.value, lb_price_20$p.value)
)
print(ljung_box_summary)

par(mfrow = c(1, 2))
acf(residuals(fit_r_000), main = "ACF: Returns Model Residuals", lag.max = 20)
acf(residuals(fit_p_010), main = "ACF: Price Model Residuals", lag.max = 20)
par(mfrow = c(1, 1))
```

```{r residual-squared-check}
lb_returns_sq <- Box.test(residuals(fit_r_000)^2, lag = 10, type = "Ljung-Box", fitdf = 0)
print(lb_returns_sq)
```

## Residual Diagnostics

The AICc-selected models for each branch - ARIMA(0,0,0) for returns and
ARIMA(0,1,0) for price (both random-walk-family baselines) - were checked for
residual white noise using the Ljung-Box test at lags 10 and 20.

| Model | Lag | p-value |
|---|---|---|
| Returns ARIMA(0,0,0) | 10 | 0.0073 |
| Returns ARIMA(0,0,0) | 20 | 0.0183 |
| Price ARIMA(0,1,0) | 10 | 0.0023 |
| Price ARIMA(0,1,0) | 20 | 0.0027 |

Both models formally reject the white-noise null at the 5% level, indicating
some residual autocorrelation remains. However, the residual ACF plots show
this structure is weak in magnitude - almost all lags sit inside the ±0.088
significance bounds, with only isolated small spikes near lags 7, 8, 10, and
17 (the same lags flagged during order identification), consistent with a
marginal rather than substantial violation.

A stronger diagnostic separates mean-structure from variance-structure
violations directly: applying the Ljung-Box test to the **squared** residuals
of the returns model gives X-squared = 74.81, p = 5.18e-12 - overwhelmingly
stronger evidence than the raw-residual test. This confirms the residual
failure is driven predominantly by volatility clustering (the same
conditional heteroskedasticity identified in the EDA and flagged as a
contributor to KPSS's borderline rejection in the Stationarity section), not
by unmodelled linear dependence in the mean. AR/MA/ARIMA models are, by
construction, unable to represent this kind of variance structure - they
model only the conditional mean.

**Conclusion:** the AICc-selected random-walk-family models are an honest,
well-justified fit to the mean structure of this data (consistent with the
near-market-efficiency expected of daily equity returns), but leave residual
volatility clustering unexplained. This is treated as a genuine limitation of
the AR/MA/ARIMA model class for this dataset, not a fitting error, and is
returned to explicitly in the Discussion section - GARCH-family models are
the natural extension for capturing this structure, primarily for improved
prediction interval calibration rather than point-forecast accuracy, given
the efficient-market constraint on predictability of the conditional mean
itself.

<!--
TODO: revisit this section after CV and final evaluation - check whether the
weak residual autocorrelation at lags 7/8/10/17 has any measurable effect on
out-of-sample forecast accuracy at any horizon, or remains purely a
mean-structure curiosity with no practical forecasting consequence.
-->
## Cross-Validation

## Final Evaluation

## Discussion
