---
title: "Financial Forecasting: AR/MA/ARIMA Model Comparison on AAPL - FRNBEN003"
format: html
---
## Forecast Horizon and Justification

## Forecast Horizon and Justification

This project forecasts AAPL daily closing price and log returns over three
horizons: 1 trading day, 5 trading days (one week), and 20 trading days
(approximately one month), rather than a single fixed horizon. This choice
was deliberate, and the results below confirm it was the right one: the three
horizons do not tell the same story, and a single-horizon comparison would
have missed that entirely.

Financial return series are widely documented to behave close to a random
walk: returns show near-zero linear autocorrelation at almost any lag,
meaning today's return carries very little information about tomorrow's
direction. A natural expectation following from this is that any genuine
short-range dependence a model captures would be most exploitable at short
horizons, with relative advantage over a naive benchmark shrinking as the
horizon lengthens and forecast uncertainty compounds. As the results below
show, this expectation held for the price branch (accuracy degraded
monotonically with horizon) but not for the returns branch, where MASE
relative to naive actually *improved* with horizon, for reasons tied to how
the naive benchmark itself behaves at longer horizons, not to the fitted
model becoming more accurate in absolute terms. This divergence between
branches is itself one of the project's findings, discussed below.

The three horizons were compared using the AICc-selected model in each
branch — ARIMA(0,0,0) for log returns, ARIMA(0,1,0) for price, via
rolling-origin cross-validation on the training set and, finally, on the
untouched test set with returns converted to an implied price forecast for a
fair dollar-scale comparison. The pattern that emerged was more nuanced than
simple "accuracy degrades with horizon" and differed by branch:

- **Price branch**: RMSE grew sharply and monotonically with horizon
  (0.68 → 1.68 → 3.87 in cross-validation; a similar pattern on the test
  set), as expected — the model's forecast stays close to the last observed
  price while the actual price continues to move, so error compounds with
  time.
- **Returns branch**: RMSE stayed roughly flat across horizons, since the
  fitted model's forecast (essentially the mean return) does not vary with
  horizon. MASE, however, *improved* at longer horizons (from ~0.98 at 1-day
  to ~0.76 at 20-day) — not because the model became more accurate in
  absolute terms, but because the one-step-ahead naive benchmark used to
  scale MASE degrades faster with horizon than the near-constant fitted
  forecast does.
- **Cross-branch, test set**: converting the returns-branch forecast to an
  implied price outperformed the price-branch forecast directly at every
  horizon, with the margin growing from roughly 1% at 1-day to roughly 13% at 20-day,
  consistent with geometric (percentage-based) compounding being more
  appropriate than arithmetic compounding as horizon increases.

Practically, this means near-term forecasting (1-day) is where the branches
are most similar and where any genuine short-range predictability would be
most exploitable, while the medium and longer horizons (5-day, 20-day) are
where the choice of modelling scale, price versus returns, has the largest
practical impact on forecast accuracy, and where the models' shared inability
to capture volatility clustering (see Residual Diagnostics and Discussion)
is most consequential for real forecast uncertainty. Comparing all three
horizons, rather than reporting a single number, was therefore essential to
surfacing this finding rather than obscuring it.


## EDA

```{r setup}
set.seed(2026)
library(readr)
library(dplyr)
library(tsibble)
library(feasts)
library(ggplot2)
library(here)
```

```{r load data}
TICKER <- "AAPL"
start_date <- as.Date("2017-01-01")
end_date <- as.Date("2019-12-31")
Train_frac <- 0.66

raw <- read_csv(
  here::here("data", "sp500_stocks.csv"),
  col_types = cols(
    date   = col_date(format = "%Y-%m-%d"),
    open   = col_double(),
    high   = col_double(),
    low    = col_double(),
    close  = col_double(),
    volume = col_double(),
    symbol = col_character()
  )
)

stock_window <- raw %>%
  filter(symbol == TICKER,
         date >= start_date,
         date <= end_date) %>%
  arrange(date) %>% 
  mutate(trading_day = row_number()) 

n_total <- nrow(stock_window)
n_train <- floor(n_total * Train_frac)

train <- stock_window %>% filter(trading_day <= n_train)
test <- stock_window %>% filter(trading_day > n_train)

cat("Total obs:", n_total, "\n")
cat("Train:", nrow(train), "|", as.character(min(train$date)), "to", as.character(max(train$date)), "\n")
cat("Test: ", nrow(test),  "|", as.character(min(test$date)),  "to", as.character(max(test$date)), "\n")

```

```{r eda-plots}
train <- train %>%
  mutate(log_close = log(close),
         log_return = log_close - lag(log_close))

# Price Plot
p_price <- ggplot(train, aes(x = date, y = close)) +
    geom_line(color = "steelblue") +
    labs(title = paste(TICKER, "- Training Set Closing Price"),
    x = "Date", y = "Close Price ($)") +
    scale_x_date(date_breaks = "3 months", date_labels = "%b %Y")


p_price

p_returns <- ggplot(train %>% filter(!is.na(log_return)), aes(x = date, y = log_return)) +
    geom_line(color = "darkorange", linewidth = 0.3) +
    geom_hline(yintercept = 0, color = "grey40", linetype = "dashed") +
    labs(title = paste(TICKER, "- Training Set Log Returns"),
    x = "Date", y = "Log Return") +
    scale_x_date(date_breaks = "3 months", date_labels = "%b %Y")
p_returns

```

```{r eda-distribution}
returns_clean <- train %>% filter(!is.na(log_return))

summary_stats <- data.frame(
    Series = c("Close Price", "Log Return"),
    Mean = c(mean(train$close), mean(returns_clean$log_return)),
    SD = c(sd(train$close), sd(returns_clean$log_return)),
    Skewness = c(
        mean((train$close - mean(train$close))^3) / sd(train$close)^3,
        mean((returns_clean$log_return - mean(returns_clean$log_return))^3) / sd(returns_clean$log_return)^3
    ),
    Kurtosis = c(
        mean((train$close - mean(train$close))^4) / sd(train$close)^4,
        mean((returns_clean$log_return - mean(returns_clean$log_return))^4) / sd(returns_clean$log_return)^4
    )
)

print(summary_stats)

# Histogram of returns vs normal
Hist_returns <- ggplot(returns_clean, aes(x = log_return)) +
    geom_histogram(aes(y = after_stat(density)), bins = 40, fill = "steelblue", alpha = 0.6) +
    stat_function(fun = dnorm, args = list(mean = mean(returns_clean$log_return), sd = sd(returns_clean$log_return)), color = "red", linewidth = 1) +
    labs(title = paste(TICKER, "- Log Returns Distribution vs Normal"),
         x = "Log Return", y = "Density")
Hist_returns
#QQ - plot
QQ_returns <- ggplot(returns_clean, aes(sample = log_return)) +
    stat_qq() +
    stat_qq_line(color = "red") +
    labs(title = paste(TICKER, "- Log Returns Q-Q Plot"),
         x = "Theoretical Quantiles", y = "Sample Quantiles")
QQ_returns
```

```{r eda-acf}
returns_vec <- returns_clean$log_return

par(mfrow = c(3, 2))

acf(train$close, main = "ACF: Price", lag.max = 40)
pacf(train$close, main = "PACF: Price", lag.max = 40)

acf(returns_vec, main = "ACF: Returns", lag.max = 40)
pacf(returns_vec, main = "PACF: Returns", lag.max = 40)

acf(returns_vec^2, main = "ACF: Squared Returns", lag.max = 40)
pacf(returns_vec^2, main = "PACF: Squared Returns", lag.max = 40)

par(mfrow = c(1, 1))

```

## Stationarity
```{r stationarity-test}
library(tseries)

adf_price <- adf.test(train$close)
kpps_price <- kpss.test(train$close, null = "Level")

adf_returns <- adf.test(returns_vec)
kpps_returns <- kpss.test(returns_vec, null = "Level")

stationarity_results <- data.frame(
    Series = c("Close Price", "Close Price", "Log Return", "Log Return"),
    Test = c("ADF", "KPSS", "ADF", "KPSS"),
    Statistic = c(adf_price$statistic, kpps_price$statistic, adf_returns$statistic, kpps_returns$statistic),
    P_Value = c(adf_price$p.value, kpps_price$p.value, adf_returns$p.value, kpps_returns$p.value)
)

print(stationarity_results)


```

## Stationarity

Formal stationarity testing was carried out using the ADF and KPSS tests, which
have opposite null hypotheses and are read together for a more reliable
conclusion (ADF's null is a unit root/non-stationarity; KPSS's null is
stationarity).

| Series | Test | Statistic | p-value |
|---|---|---|---|
| Close Price | ADF | -0.793 | 0.962 |
| Close Price | KPSS | 6.856 | 0.010 |
| Log Return | ADF | -7.565 | 0.010 |
| Log Return | KPSS | 0.551 | 0.030 |

**Price**: ADF fails to reject the unit-root null (p = 0.962) and KPSS rejects
stationarity (p = 0.010). Both tests agree: price is non-stationary. This
confirms the visual evidence from the EDA (persistent upward trend, ACF
decaying extremely slowly even at lag 40) and the PACF's sharp cutoff after
lag 1, consistent with a random-walk-type process. Price requires differencing
(d = 1) before AR/MA-type modelling.

**Log returns**: ADF strongly rejects the unit-root null (p = 0.010), while
KPSS's rejection (p = 0.030) is borderline and most plausibly attributable to
the conditional heteroskedasticity (volatility clustering) already identified
in the EDA — via the widened swings visible in the returns plot, excess
kurtosis (5.92), and the more persistent squared-returns ACF — rather than
genuine non-stationarity in the mean. KPSS is known to be sensitive to this
kind of heteroskedasticity even when the mean is stationary. Log returns are
therefore treated as stationary (d = 0) for AR/MA/ARMA modelling.

**Decision going forward:** price will be modelled via ARIMA(p,1,q) (differencing
handled internally); log returns will be modelled via ARMA(p,q) (already
stationary, no further differencing). Because these two branches use different
orders of differencing, AICc cannot be used to compare across them directly —
only within each branch. Cross-branch comparison (price-based vs returns-based
forecasts) will instead rely on out-of-sample RMSE from the test set, which
remains valid regardless of differencing order.

<!--
TODO: revisit this section once order identification and model fitting are
done - check this interpretation still holds and reference the final chosen
models here if useful.
-->

## ACF/PACF Identification
``` {r order-id-acf-pacf}
par(mfrow = c(1, 2))
acf(returns_vec, main = "ACF: Log Returns (training)", lag.max = 20)
pacf(returns_vec, main = "PACF: Log Returns (training)", lag.max = 20)

par(mfrow = c(1, 1))

n <- length(returns_vec)
ci <- 1.96/sqrt(n)
acf_vals <- acf(returns_vec, lag.max = 20, plot = FALSE)
pacf_vals <- pacf(returns_vec, lag.max = 20, plot = FALSE)

cat("95% significance bounds: +/-", round(ci,4), "\n")
cat("Significant ACF lags (excl. lag 0):", which(abs(acf_vals$acf[-1]) > ci), "\n")
cat("Significant PACF lags:", which(abs(pacf_vals$acf) > ci), "\n")


diff_price <- diff(train$close)

par(mfrow = c(1, 2))
acf(diff_price, main = "ACF: Differenced Price (training)", lag.max = 20)
pacf(diff_price, main = "PACF: Differenced Price (training)", lag.max = 20)
par(mfrow = c(1, 1))

n_d <- length(diff_price)
ci_d <- 1.96/sqrt(n_d)
acf_vals_d <- acf(diff_price, lag.max = 20, plot = FALSE)
pacf_vals_d <- pacf(diff_price, lag.max = 20, plot = FALSE)

cat("95% significance bounds (diff price): +/-", round(ci_d,4), "\n")
cat("Significant ACF lags (excl. lag 0):", which(abs(acf_vals_d$acf[-1]) > ci_d), "\n")
cat("Significant PACF lags (diff price):", which(abs(pacf_vals_d$acf) > ci_d), "\n")
```


## ACF/PACF Identification

Order identification was carried out on the returns series only (already
established as stationary), using ACF and PACF computed to lag 20, with
significance assessed numerically against the ±1.96/√n critical bound rather
than by visual inspection alone.

Lags 1–6 fall comfortably within the significance bounds in both ACF and PACF,
showing no early cutoff or gradual decay pattern, consistent with the
near-market-efficiency expected of daily equity returns (little to no linear
autocorrelation at short lags). Isolated significant spikes appear at lags 7,
8, 10, and 17, flagged identically in both ACF and PACF. This does not match
either the textbook AR signature (PACF cutoff, ACF decay) or MA signature (ACF
cutoff, PACF decay), and is most plausibly explained by multiple-testing noise
(with 20 lags tested at a 95% threshold, roughly 1 spurious spike is expected
by chance) rather than genuine structure, particularly as the flagged lags
have no clear interpretable spacing (e.g. no weekly/5-day pattern). This
reading was later corroborated by the residual diagnostics: the raw-residual
Ljung-Box test was only borderline significant (driven by these same lags),
while the squared-residual test was overwhelming (p = 5.18e-12), indicating
the mean-structure signal at these lags is genuinely weak, and the dominant
source of model inadequacy is volatility clustering rather than an
unmodelled AR/MA component at lags 7/8/10/17.

**Decision:** given the weak and ambiguous short-lag signal, small candidate
orders are fitted and compared via AICc and residual diagnostics, rather than
reading a specific (p, q) directly off the plot. This is a deliberate choice
given the data, not a shortcut, forcing a larger/more specific order onto an
ambiguous ACF/PACF read would not be justifiable.

**Candidates — Returns branch (ARMA, d = 0):**

| Model | AICc |
|---|---|
| **ARIMA(0,0,0)** | **-2772.501** |
| AR(1) | -2772.305 |
| MA(1) | -2772.420 |
| ARMA(1,1) | -2770.538 |

ARIMA(0,0,0) has the lowest (best) AICc, though the margin over AR(1) and
MA(1) is small (< 0.2) and not practically distinguishable; ARMA(1,1) is the
clearest laggard (~2 points worse). ARIMA(0,0,0) was carried forward as the
returns-branch model for cross-validation and final evaluation.

**Price branch identification.** Order identification was repeated on the
first-differenced raw close price (not log price, to keep this branch
conceptually distinct from the returns branch rather than redundant with it -
differencing log price would be mathematically equivalent to log returns).

The differenced price ACF/PACF show the same pattern as the returns branch:
lags 1-6 within the significance bounds in both plots, with isolated
significant spikes at lags 7, 8, 10, and 17, flagged identically in ACF and
PACF (95% bound = ±0.088). This near-identical structure between the two
branches is itself a notable finding: despite deliberately using raw price
differences rather than log returns, the two series behave very similarly.
This is consistent with the standard result that log returns closely
approximate percentage price changes when period-to-period moves are small
relative to the price level, as is typical for daily equity data, the
distinction between "dollar change" and "log/percentage change" matters much
less when moves are a small fraction of price. Despite this structural
similarity, the two branches' final test-set forecasting performance
diverged materially (see Final Evaluation and Discussion), the resemblance
in ACF/PACF shape does not imply the two branches forecast equally well once
compounded over multiple steps.

**Candidates - Price branch (ARIMA, d = 1):**

| Model | AICc |
|---|---|
| **ARIMA(0,1,0)** | **937.003** |
| ARIMA(0,1,1) | 937.433 |
| ARIMA(1,1,0) | 937.580 |
| ARIMA(1,1,1) | 939.108 |

As with the returns branch, ARIMA(0,1,0) wins by a small margin over the
single-parameter alternatives, with ARIMA(1,1,1) the clearest laggard.
ARIMA(0,1,0) was carried forward as the price-branch model for
cross-validation and final evaluation. Note that AICc values are not
comparable between the returns and price tables above, as the two branches
differ in order of differencing (d); cross-branch comparison was instead
carried out via out-of-sample RMSE in the Final Evaluation section.

## Model Fitting

```{r fit-returns}
aicc <- function(fit, n_eff){
    k <- length(fit$coef) + 1 #+1 for sigma^2
    fit$aic + (2 * k * (k+1))/(n_eff - k - 1)

}

fit_r_000 <- arima(returns_vec, order = c(0, 0, 0))
fit_r_100 <- arima(returns_vec, order = c(1, 0, 0))
fit_r_001 <- arima(returns_vec, order = c(0, 0, 1))
fit_r_101 <- arima(returns_vec, order = c(1, 0, 1))

n_r <- length(returns_vec)
returns_results <- data.frame(
    Model = c("ARIMA(0,0,0)", "AR(1)", "MA(1)", "ARMA(1,1)"),
    AICc = c(aicc(fit_r_000, n_r), aicc(fit_r_100, n_r), aicc(fit_r_001, n_r), aicc(fit_r_101, n_r))
)

print(returns_results)
```

```{r fit-price}
fit_p_010 <- arima(train$close, order = c(0, 1, 0))
fit_p_110 <- arima(train$close, order = c(1, 1, 0))
fit_p_011 <- arima(train$close, order = c(0, 1, 1))
fit_p_111 <- arima(train$close, order = c(1, 1, 1))

n_p <- length(train$close) - 1
price_results <- data.frame(
    Model = c("ARIMA(0,1,0)", "ARIMA(1,1,0)", "ARIMA(0,1,1)", "ARIMA(1,1,1)"),
    AICc = c(aicc(fit_p_010, n_p), aicc(fit_p_110, n_p), aicc(fit_p_011, n_p), aicc(fit_p_111, n_p))
)

print(price_results)
```


```{r residual-diagnostics}
lb_returns <- Box.test(residuals(fit_r_000), lag = 20, type = "Ljung-Box", fitdf = 0)
lb_price <- Box.test(residuals(fit_p_010), lag = 20, type = "Ljung-Box", fitdf = 0)

cat("Returns model = ARIMA(0,0,0):\n")
print(lb_returns)
cat("\nPrice model = ARIMA(0,1,0):\n")
print(lb_price)
```

```{r residual-diagnostics-compare}
lb_returns_10 <- Box.test(residuals(fit_r_000), lag = 10, type = "Ljung-Box", fitdf = 0)
lb_returns_20 <- Box.test(residuals(fit_r_000), lag = 20, type = "Ljung-Box", fitdf = 0)
lb_price_10   <- Box.test(residuals(fit_p_010), lag = 10, type = "Ljung-Box", fitdf = 0)
lb_price_20   <- Box.test(residuals(fit_p_010), lag = 20, type = "Ljung-Box", fitdf = 0)

ljung_box_summary <- data.frame(
  Model = rep(c("Returns ARIMA(0,0,0)", "Price ARIMA(0,1,0)"), each = 2),
  Lag = c(10, 20, 10, 20),
  p_value = c(lb_returns_10$p.value, lb_returns_20$p.value,
              lb_price_10$p.value, lb_price_20$p.value)
)
print(ljung_box_summary)

par(mfrow = c(1, 2))
acf(residuals(fit_r_000), main = "ACF: Returns Model Residuals", lag.max = 20)
acf(residuals(fit_p_010), main = "ACF: Price Model Residuals", lag.max = 20)
par(mfrow = c(1, 1))
```

```{r residual-squared-check}
lb_returns_sq <- Box.test(residuals(fit_r_000)^2, lag = 10, type = "Ljung-Box", fitdf = 0)
print(lb_returns_sq)
```

## Residual Diagnostics

The AICc-selected models for each branch, ARIMA(0,0,0) for returns and
ARIMA(0,1,0) for price (both random-walk-family baselines), were checked for
residual white noise using the Ljung-Box test at lags 10 and 20.

| Model | Lag | p-value |
|---|---|---|
| Returns ARIMA(0,0,0) | 10 | 0.0073 |
| Returns ARIMA(0,0,0) | 20 | 0.0183 |
| Price ARIMA(0,1,0) | 10 | 0.0023 |
| Price ARIMA(0,1,0) | 20 | 0.0027 |

Both models formally reject the white-noise null at the 5% level, indicating
some residual autocorrelation remains. However, the residual ACF plots show
this structure is weak in magnitude, almost all lags sit inside the ±0.088
significance bounds, with only isolated small spikes near lags 7, 8, 10, and
17 (the same lags flagged during order identification), consistent with a
marginal rather than substantial violation.

A stronger diagnostic separates mean-structure from variance-structure
violations directly: applying the Ljung-Box test to the **squared** residuals
of the returns model gives X-squared = 74.81, p = 5.18e-12, overwhelmingly
stronger evidence than the raw-residual test. This confirms the residual
failure is driven predominantly by volatility clustering (the same
conditional heteroskedasticity identified in the EDA and flagged as a
contributor to KPSS's borderline rejection in the Stationarity section), not
by unmodelled linear dependence in the mean. AR/MA/ARIMA models are, by
construction, unable to represent this kind of variance structure, they
model only the conditional mean.

**Conclusion:** the AICc-selected random-walk-family models are an honest,
well-justified fit to the mean structure of this data (consistent with the
near-market-efficiency expected of daily equity returns), but leave residual
volatility clustering unexplained. This is treated as a genuine limitation of
the AR/MA/ARIMA model class for this dataset, not a fitting error, and is
returned to explicitly in the Discussion section, GARCH-family models are
the natural extension for capturing this structure, primarily for improved
prediction interval calibration rather than point-forecast accuracy, given
the efficient-market constraint on predictability of the conditional mean
itself.

The cross-validation and final evaluation results (see those sections)
provide indirect evidence on whether the weak residual autocorrelation at
lags 7, 8, 10, and 17 carries any practical forecasting consequence: AR(1),
MA(1), and ARMA(1,1), the candidates best placed to capture any leftover
short-lag mean structure, performed no better than the ARIMA(0,0,0)/
ARIMA(0,1,0) baselines in either cross-validation or on the test set, in
either branch. No model specifically targeting lags 7/8/10/17 was fitted, as
doing so without a principled justification would risk fitting noise rather
than genuine structure (see ACF/PACF Identification). Taken together, this
supports treating the residual autocorrelation as a marginal, low-consequence
curiosity for point forecasting, consistent with the interpretation above
that volatility clustering, not unmodelled mean structure, is the
substantive limitation of this model class for this dataset.
## Cross-Validation

```{r cv-setup}
source(here::here("projects", "project-01-financial-forecasting", "R", "cv_functions.R"))

```

```{r cv-returns}

returns_orders <- list("ARIMA(0,0,0)" = c(0,0,0), "AR(1)" = c(1,0,0),
                        "MA(1)" = c(0,0,1), "ARMA(1,1)" = c(1,0,1))

cv_returns_results <- data.frame()
for (nm in names(returns_orders)) {
  res <- rolling_cv(returns_vec, returns_orders[[nm]], horizons = c(1,5,20),
                     initial = 300, step = 5)
  res$Model <- nm
  cv_returns_results <- rbind(cv_returns_results, res)
}
print(cv_returns_results[, c("Model","Horizon","RMSE","MAE","MASE","n_origins")])


cat("SD of returns_vec:", sd(returns_vec), "\n")
cat("Q (mean abs diff of returns_vec):", mean(abs(diff(returns_vec))), "\n")
cat("Summary of returns_vec:\n")
print(summary(returns_vec))
cat("Summary of diff(returns_vec):\n")
print(summary(diff(returns_vec)))
```


``` {r cv-price}

price_orders <- list("ARIMA(0,1,0)" = c(0,1,0), "ARIMA(1,1,0)" = c(1,1,0),
                     "ARIMA(0,1,1)" = c(0,1,1), "ARIMA(1,1,1)" = c(1,1,1))

cv_price_results <- data.frame()
for (nm in names(price_orders)) {
  res <- rolling_cv(train$close, price_orders[[nm]], horizons = c(1,5,20),
                     initial = 300, step = 5)
  res$Model <- nm
  cv_price_results <- rbind(cv_price_results, res)
}
print(cv_price_results[, c("Model","Horizon","RMSE","MAE","MASE","n_origins")])

```

## Final Evaluation

```{r final-eval-setup}
source(here::here("projects", "project-01-financial-forecasting", "R", "forecast_functions.R"))

full_data <- stock_window
full_price <- full_data$close
full_returns <- c(NA, diff(log(full_data$close)))

test_start <- nrow(train)
test_end <- nrow(full_data)

```

``` {r final-eval-run}
final_price_results <- final_eval_price(full_price, order = c(0,1,0), horizons = c(1,5,20), test_start = test_start, test_end = test_end)
final_price_results$Branch <- "Price ARIMA(0,1,0)"

final_returns_results <- final_eval_returns_as_price(full_returns,full_price, order = c(0,0,0), horizons = c(1,5,20), test_start = test_start, test_end = test_end)
final_returns_results$Branch <- "Returns ARIMA(0,0,0)"

final_comparison <- rbind(final_price_results, final_returns_results)
print(final_comparison[, c("Branch","Horizon","RMSE","n_origins")])
```

## Discussion

### Summary of findings

This project set out to compare AR, MA, and ARIMA models for forecasting AAPL
daily closing prices, modelled both directly (price, ARIMA branch, d = 1) and
via log returns (ARMA branch, d = 0). Every stage of the analysis pointed in
the same direction.

The EDA identified a clear upward price trend interrupted by a sharp late-2018
decline, near-zero linear autocorrelation in returns, and visually apparent
volatility clustering concentrated around that decline. Formal stationarity
testing confirmed price is non-stationary (ADF/KPSS in agreement) and returns
are stationary in the mean, with KPSS's borderline rejection attributed to
conditional heteroskedasticity rather than a genuine unit root. Order
identification via ACF/PACF showed no clean AR or MA signature at short lags
in either branch - isolated spikes at lags 7, 8, 10, and 17 were judged more
consistent with multiple-testing noise than genuine structure, given their
lack of interpretable spacing.

This ambiguity was confirmed empirically: AICc favoured the simplest possible
model in both branches, ARIMA(0,0,0) for returns and ARIMA(0,1,0) for price, with AR(1), MA(1), and ARMA(1,1)-type extensions offering no meaningful
improvement in either branch. Rolling-origin cross-validation on the training
set reproduced this ranking out-of-sample, with MASE below 1 (returns) and
above 1 (price) across all horizons tested. On the untouched test set, the
returns-branch forecast, converted to an implied price via geometric
compounding for direct dollar-scale comparability - outperformed the
price-branch forecast at every horizon, with the margin growing from ~1% at
1-day to ~13% at 20-day, consistent with geometric random-walk compounding
being more appropriate than an arithmetic one over longer horizons.

Residual diagnostics provided the clearest explanation for why no candidate
model performed strongly in an absolute sense. Ljung-Box tests on the raw
residuals showed only a weak, borderline violation of white noise, but the
same test applied to **squared** residuals was overwhelming
(X-squared = 74.81, p = 5.18e-12), confirming that the volatility clustering
identified in the EDA persists, unmodelled, in the fitted models' residuals.
This is not a fitting error: AR/MA/ARIMA models are constructed to capture
linear dependence in the conditional mean only, and have no mechanism to
represent time-varying variance.

### What this means

Taken together, these results are best read not as a failed search for a
good forecasting model, but as empirical evidence that AAPL's daily returns
over this window behave close to the near-market-efficient random walk that
financial theory predicts: little linear structure exists in the mean for a
model of this class to find, and the AICc/CV/test-set procedures correctly
converged on the simplest model available rather than overfitting to noise
in pursuit of a more impressive-looking result. The one place a clear,
well-evidenced departure from "nothing to model" was found, volatility
clustering, is precisely the structure this model class cannot represent.

### Limitations

- The price-branch and returns-branch winning models are not fully
  equivalent despite their conceptual similarity: ARIMA(0,0,0) on returns
  implicitly estimates a mean/drift term that compounds forward under the
  implied-price conversion, whereas ARIMA(0,1,0) on price is fit without
  drift by default. Part of the returns branch's test-set advantage may
  therefore reflect this drift rather than the arithmetic-vs-geometric
  distinction alone; a drift-included price model was identified as a useful
  follow-up check but not yet completed.
- Small-sample convergence warnings occurred for ARMA(1,1) at early
  cross-validation origins with limited training data; affected origins were
  excluded from that model's error calculation rather than treated as fatal.
- The 3-year training/test window (2017-2019) does not include the training
  set volatility regime experienced in the test period; results may not
  generalise to periods of substantially different market conditions.

### Future work

The residual diagnostics motivate a direct and well-justified next step:
GARCH-family models, planned as a subsequent project in this portfolio.
Unlike the models tested here, GARCH explicitly models time-varying
conditional variance, directly targeting the volatility clustering this
project identified but could not represent. Given the efficient-market
constraint already evident in this project's point-forecast results, GARCH
is expected to offer limited improvement in point-forecast accuracy, but
should meaningfully improve prediction interval calibration - narrow,
constant-width intervals from this project's ARIMA-family models are likely
to understate real forecast uncertainty, particularly around volatile
periods such as the Q4 2018 decline identified in the EDA.

<!--
TODO: revisit once the drift-included price model (ARIMA(0,1,0) with
include.mean/drift) is fitted, to confirm whether the returns-branch
advantage survives after controlling for drift.
-->
